---
title: "Plant Community Response"
output:
  pdf_document: default
  html_document: default
date: "2025-08-05"
---

```{r setup, include=TRUE}
#Load libraries
library(readxl)
library(ggplot2)
library(dplyr)
library(viridis)
library(ggpubr)
library(rstatix)
library(glmmTMB)
library(boot)
library(tidyr)
library(vegan)
library(tibble)  
library(dplyr)   
library(purrr)   
library(boot)  
```

# Load the dataset and calcualte relative biomass
To calculate relative biomass, the total biomass per species was divided by the
total biomass within a mesocosm. 

```{r}
#Import the dataset
data_mortality <- read_excel(file.path(
  "C:/Users/Joyalea/Documents/UBCO/Thesis",
  "Final_Files",
  "RQ3_PlantCommunityResponse",
  "MasterCopy_Biomass_2024.xlsx"
))

#Add treatment group column
data_mortality <- data_mortality %>%
  mutate(treatment = ifelse(mesocosm_id %in% c(1:7), "Control",
                            ifelse(mesocosm_id %in% c(8:14), "197198",
                                   ifelse(mesocosm_id %in% c(15:21), "240448",
                            ifelse(mesocosm_id %in% c(22:28), "240720", NA)))),
         block = ifelse(mesocosm_id %in% c(15, 20, 22, 12, 11, 14, 24, 27, 
                                           19, 21, 25, 3, 2, 7), "A",
                        ifelse(mesocosm_id %in% c(4, 17, 9, 28, 6, 1, 5, 18, 8, 
                                              26, 16, 23, 13, 10), "B", NA)))
#Review the data
head(data_mortality)
str(data_mortality)

#Summarize count of individuals
counts_df <- data_mortality %>%
  filter(species == "Bromus", is.na(biomass)) %>%
  group_by(species, treatment) %>%
  summarise(n_individuals = n(), .groups = "drop")

#Summarize relative biomass (species biomass / total meso biomass)
rel_biomass <- data_mortality %>%
  group_by(mesocosm_id) %>%
  mutate(total_biomass = sum(biomass, na.rm = TRUE)) %>%
  group_by(mesocosm_id, species, treatment) %>%
  summarise(
    species_biomass = sum(biomass, na.rm = TRUE),
    total_biomass = unique(total_biomass),
    rel_biomass = species_biomass / total_biomass,
    .groups = "drop"
  )

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Analyse changes in plant community structure (beta diversity) using relative host biomass
Relative host biomass was used as opposed to counts to account for overall
growth of a species in a mesocosm, which can indirectly account for mortality

```{r}

#Change data structure for bray analysis (convert columns to rows)
pivot_rel_biomass <- rel_biomass %>%
  select(mesocosm_id, species, rel_biomass) %>%
  pivot_wider(names_from = species, values_from = rel_biomass, values_fill = 0) %>%
  column_to_rownames("mesocosm_id")

#Create metadata
metadata <- rel_biomass %>%
  select(mesocosm_id, treatment) %>%
  distinct()

metadata <- metadata %>%
  filter(mesocosm_id %in% rownames(pivot_rel_biomass)) %>%
  arrange(match(mesocosm_id, rownames(pivot_rel_biomass)))

#Check it worked correctly
View(metadata)

#Create Bray-Curtis dissimilarity matrix
bray_dist <- vegdist(pivot_rel_biomass, method = "bray")

print(adonis2(bray_dist ~ treatment, data = metadata, permutations = 999))
#Significant

#Check dispersion to ensure ANOVA assumptions are met
dispersion <- betadisper(bray_dist, metadata$treatment)
anova(dispersion) #Not significant
plot(dispersion)
boxplot(dispersion)

#Pairwise comparisons
pairwiseAdonis::pairwise.adonis(
  x = as.matrix(bray_dist),
  factors = metadata$treatment,
  sim.function = "vegdist",
  sim.method = "bray",
  p.adjust.m = "BH"
)

#SIMPER analysis to see which species are driving the difference
simper_results <- simper(pivot_rel_biomass, group = metadata$treatment, 
                         permutations = 999)

#View SIMPER results for each treatment pair
summary(simper_results) #Driven by Gaillardia


```

## Visualise relative biomass by treatment using a stacked barplot

```{r}

#Calculate mean +/- SD of relative biomass per species per treatment
rel_biomass <- rel_biomass %>%
  group_by(treatment, species) %>%
  summarise(mean_rel_biomass = mean(rel_biomass, na.rm = TRUE),
            sd_rel_biomass = sd(rel_biomass, na.rm = TRUE),
            .groups = "drop")

#Plot: stacked bar of mean relative biomass per treatment
rel_biomass_plot<-ggplot(rel_biomass, aes(x = treatment, y = mean_rel_biomass, fill = species)) +
  geom_col(position = "stack", width = 0.8, alpha = 0.95) +
  labs(title = "Mean Relative Biomass Composition by Treatment",
       y = "Relative Biomass",
       x = "Treatment") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1), expand = c(0, 0)) +
  scale_fill_brewer(palette = "Paired", name = "Species") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )
print(rel_biomass_plot)

```

## Evaluate alpha diversity metrics

```{r}

#Evaluate community richness (# of species present in a mesocosm)
richness <- specnumber(pivot_rel_biomass) #All species present in all mesocosms

#Shannon diversity
shannon_div <- diversity(pivot_rel_biomass, index = "shannon")

#Pielou’s Evenness
evenness <- shannon_div / log(richness)

#Combine into data frame
diversity_df <- data.frame(
  mesocosm_id = rownames(pivot_rel_biomass),
  shannon = shannon_div,
  richness = richness,
  evenness = evenness
) %>%
  # Ensure mesocosm_id is character for joining
  mutate(mesocosm_id = as.factor(mesocosm_id)) %>%
  # Add treatment or other metadata
  left_join(metadata %>% mutate(mesocosm_id = as.factor(mesocosm_id)), 
            by = "mesocosm_id")

#Test for treatment effects on evenness
kruskal_result <- kruskal.test(evenness ~ treatment, data = diversity_df)
kruskal_result

#Test for treatment effects on shannon diversity
#Test for treatment effects on evenness
kruskal_result_shannon <- kruskal.test(shannon_div ~ treatment, data = diversity_df)
kruskal_result_shannon

```

## Analysis of host mortality 

```{r, echo=FALSE}

#Plot host mortality based on the count_df dataframe
print(ggplot(counts_df, aes(x = species, y = n_individuals, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7, alpha = 0.9) +
  labs(title = "Number of Individuals by Species and Treatment",
       y = "Number of Mortalities") +
  scale_fill_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  ))

#Convert to % survivial
survivors_df <- data_mortality %>%
  filter(!is.na(species), biomass > 0) %>%
  group_by(species, treatment) %>%
  summarise(n_survived = n(), .groups = "drop")

#Count total planted by group
total_df <- data_mortality %>%
  filter(!is.na(species)) %>%
  group_by(species, treatment) %>%
  summarise(n_total = n(), .groups = "drop")


#Merge and compute % survival of each host specoes
survival_df <- left_join(total_df, survivors_df, by = c("species", "treatment")) %>%
  mutate(n_survived = replace_na(n_survived, 0),  # Fill 0 where no survival recorded
         percent_survival = (n_survived / n_total) * 100)

#Plot % survival
ggplot(survival_df, aes(x = species, y = percent_survival, fill = treatment)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7, alpha = 0.9) +
  labs(title = "Percent Survival by Species and Treatment",
       y = "% Survival",
       x = "Species") +
  scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
  scale_fill_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )

#As Bromus was the only species group with variaiton in mortality. Subset Bromus.
bromus_survival <- survival_df %>%
  filter(species == "Bromus")
print(bromus_survival)

#Plot bromus % survival
bromus_percent_survival<-ggplot(bromus_survival, aes(x = treatment, y = percent_survival, fill = treatment)) +
  geom_col(width = 0.7, alpha = 0.9) +
  labs(title = "Percent Survival in Bromus by Treatment",
       y = "% Survival",
       x = "Treatment") +
  scale_y_continuous(limits = c(0, 100), expand = c(0, 0)) +
  scale_fill_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )

#Summarize deaths (biomass == NA)
deaths_df <- data_mortality %>%
  filter(species == "Bromus", is.na(biomass)) %>%
  count(treatment, name = "dead")

#Summarize survivals (biomass != NA)
survivors_df <- data_mortality %>%
  filter(species == "Bromus", !is.na(biomass)) %>%
  count(treatment, name = "alive")

#Merge and fill missing 0s
mortality_table <- full_join(deaths_df, survivors_df, by = "treatment") %>%
  mutate(across(c(dead, alive), ~replace_na(., 0))) %>%
  arrange(treatment)

#Prepare matrix to perform fisher exact test (convert table to matrix)
fisher_matrix <- mortality_table %>%
  select(dead, alive) %>%
  as.matrix()

#Fishers exact test
fisher_result <- fisher.test(fisher_matrix)
print(fisher_result) #No significant differences in bromus mortality


```
